An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-12-19 Project_0/manuscript/manuscript_R1.tex
revision for project 0 complete, save work for M-Ecov revision

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{04 December, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models treat latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other other measurement properties. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effect parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and are recommended for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

@@

As state-space models provide greater flexibility by allowing multiple
processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
al. 2018; Stock et al. 2021), one of the most immediate questions
regards the implications of mis-specification among alternative sources
of process error. Incorrect treatment of population attributes as
temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
could lead to misidentification of stock status and biased population
estimates, ultimately impacting fisheries management decisions (Legault
and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
Furthermore, biological, fishery, and observational processes are often
confounded in catch-at-age data, which may adversely affect the ability
to distinguish between true process variability and observational error
(Li et al. In press; Punt et al. 2014; Stewart and Monnahan 2017;
Cronin-Fine and Punt 2021; Fisch et al. 2023).

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model have not
been thoroughly examined or tested within a simulation-estimation
framework. Second, previous studies relied on operating models
conditioned on specific fisheries, limiting their generalizability (Li
et al. In press; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Pontavice et
al. 2022), but reliability of inferences within integrated state-space
age-structured models has not been evaluated. Evidence from other
studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016). Finally, although the importance
of autocorrelation in process errors is recognized, investigations of
the ability to distinguish state-space assessment models with and
without autocorrelation and whether such misspecification is detrimental
to estimation of important population metrics are lacking (Johnson et
al. 2016; Xu et al. 2019).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
whether OM and EM attributes affect rates of convergence and the ability
of Akaike Information Criterion (AIC) to correctly determine the source
of process error or the existence of a SRR. We also evaluate when
retrospective patterns occur and the degree of bias in outputs of the
assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Stock and Miller 2021; Miller et al. 2025).
WHAM is an R package freely available via a Github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on log-scale the random effects are additive to the
total mortality (F+M) between numbers at age, thus they modify the
survival term. However, as Stock and Miller (2021) note, these random
effects can be due to events other than mortality, such as immigration,
emigration, missreported catch, and other sources of misspecification.
For each OM, assumptions about the magnitude of the variance of process
errors and observations are required and the values we used were based
on a review of the range of estimates from Northeast Unite States (NEUS)
assessments using WHAM.

In total, we configured 72 OMs with alternative assumptions about the
source and magnitude of process errors, magnitude of observation error
in indices and age composition data, and contrast in fishing pressure
over time. For each OM, we simulated 100 population time series with
process errors and, for each time series, simulated observation data
sets. For each data set, we fitted a number of EMs that differed in
assumptions about the source of process errors, whether natural
mortality (or the median for models with process error in natural
mortality) was estimated, and whether a Beverton-Holt SRR was estimated
within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. 2025). All code
we used to perform the simulation study and summarize results can be
found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

We intended the population demographics and observation types to
represent a general NEUS groundfish stock. The population consists of 10
age classes, ages 1 to 10+, with the last being a plus group that
accumulates ages 10 and older. We assume spawning occurs annually 1/4 of
the way through the year. The maturity at age was a logistic curve with
\(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
left).

Weight at age was generated with a von Bertalanffy growth function \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_inputs_fig}, top right).

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All biological inputs to calculations of
spawning biomass per recruit (i.e., weight, maturity, and natural
mortality at age) are constant in the apparent survival (R+S)
selectivity (R+Sel), and survey catchability (R+q) process error OMs.
Therefore, steepness and unfished recruitment are also constant over the
time period for those OMs (Miller and Brooks 2021). We assumed a value
of 0.2 for the natural mortality rate in OMs without process errors on
natural mortality. We specified unfished recruitment equal to \(e^{10}\)
and \(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69
and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). For OMs with time-varying
random effects for natural mortality, steepness is not constant.
However, we used the same \(a\) and \(b\) parameters as other OMs, which
equates to a steepness and R0 at the median of the time series process
for natural mortality. Similarly, For OMs with time-varying random
effects for fishery selectivity, \Fmsy also varies temporally, so
equilibrium conditions for these OMs are defined for mean selectivity
parameters.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years followed by
fishing at \Fmsy for the last 20 years (denoted
\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
fished at \Fmsy for the entire time period (40 years). The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
the approach in Legault et al. (2023). The second scenario represents
the ideal situation where the stock is fished at an optimal level, but
provides less contrast in stock sizes over time. We specified initial
population abundance at age at the equilibrium distribution that
corresponds to fishing at either \(F = 2.5\Fmsy\) or \(F = \Fmsy\). This
implies that, for a deterministic model, the abundance at age would not
change from the first year to the next.

@@

For operating models with random effects on recruitment and(or) apparent
survival (R, R+S), we assumed marginal standard deviations for
recruitment of \(\sigma_R \in \{0.5,1.5\}\). The marginal standard
deviations for apparent survival random effects at older age classes
were \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial combination
of these process error assumptions (\(2\times 3\) levels) and scenarios
for fishing history (2 levels) and observation error (2 levels)
scenarios described above results in 24 different R
(\(\sigma_{2+} = 0\)) and R+S operating models (Table
\ref{naa_om_table}).

@@

WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
\[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\), where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters (\(a_{50}\) and slope
parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal
standard deviations we assumed for logit scale random effects were
\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
of these process error assumptions (2x2 levels) and scenarios described
above for fishing history (2 levels) and observation error (2 levels)
results in 16 different R+Sel OMs (Table \ref{sel_om_table}).

@@

We investigated the reliability of AIC-based model selection for two
purposes. First, we analyzed selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
simulated data set, we compared AIC for EMs with different process error
assumption conditional on whether median natural mortality rate and the
Beverton-Holt SRR were estimated. Second, we analyzed AIC-based
selection between EMs with and without the Beverton-Holt SRR assumed.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections indicated generally poor performance of AIC in determining
the Beverton-Holt model for a given set of OM factors (including
contrast in fishing pressure), even when the EM was configured with the
correct process error type. Therefore, we conditioned on the EMs having
the correct process error assumption and also considered the effect of
the log-standard deviation of the true log(SSB)
(\(\log \text{SD}_{\text{SSB}}\); similar to the log of the coefficient
of variation for SSB) on model selection since simulations with realized
SSB producing low and high recruitment would have larger variation in
realized SSB.

@@

We also investigated bias in estimation of various model attributes as a
measure of reliability. For a given model attribute we calculated the
relative error \begin{equation}\label{relerror}
\text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
\end{equation} from fitting a given estimating model to simulated data
set \(j\) configured for a given OM where \(\widehat \theta_j\) and
\(\theta_j\) are the estimated and true values for simulation \(j\). We
analyzed simulation results for estimates of terminal year estimates of
SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
median natural mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted
models as a measure of reliability. We calculated Mohn's \(\rho\) for
SSB, fishing mortality (averaged over all age classes), and recruitment
for each EM fit to each OM simulated data set (Mohn 1999). We fit
\(P = 7\) peels to each simulated data set and calculated Mohn's
\(\rho\) for a given attribute \(\theta\) as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(Y\) is last year of the full set of observations
and \(\widehat \theta_{y,y'}\) is the estimate for attribute \(\theta\)
in year \(y\) from a model fit using data up to year \(y'\geq y\).

@@

Because the OM and EM attributes that we investigated are numerous, we
used two methods to summarize the most important factors for differences
in results within a given OM process error type. The first method was
fitting regression models with the response being each of the measures
of reliability described above and predictor variables were defined
based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
al. 2017; Harwell et al. 2018). For the binary indicators of convergence
and AIC-based selection of a SRR, we performed logistic regressions. For
indicators of AIC-based selection of EM process error type (multiple
categories) we performed multinomial regressions. For other measures of
reliability we fit linear regression models to transformed responses.
Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the
various parameters are bounded below at -1, we used a transformation of
these values \begin{equation}\label{bias_regression_response}
y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(j\), so that values are unbounded. For relative errors, \(y_j\) is the
log-scale error. We omitted simulations where estimated attributes equal
to zero (RE = -1). For all regressions we fit separate models with just
individual OM and EM factors included, with all factors included, with
all second order interactions, and with all third order interactions.
For the multinomial regression, we used the \verb|vglm| function from
the VGAM package (Yee 2008; Yee 2015). We tabulated percent reduction in
residual deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the OM and EM attributes, and their
interactions, split the values for each measure of reliability (e.g.,
Gonzalez et al. 2018; Collier et al. 2022). We used classification trees
for categorical measures (convergence and AIC) and regression trees for
the other measures with continuous scales (relative error and Mohn's
\(\rho\)). The response variables were the same as the regressions for
the deviance reduction analyses. We used the \verb|rpart| function in
the rpart package (\url{https://cran.r-project.org/package=rpart}) to
fit trees. Full trees were determined using default settings except that
we increased the number of cross-validations to 100. For clarity, we
pruned the full trees to show just the primary branches.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Gregory L. Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alexander C. Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Gregory L.
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin
Fay\textsuperscript{4} \and Alexander C.
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{19 December, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within the International Council
for the Exploration of the Sea (ICES), Canada, and the Northeast US
(Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and
Miller 2021). State-space models treat latent population characteristics
as statistical time series with periodic observations that also may have
error due to sampling or other measurement properties. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effect parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These
random-effects approaches are considered best practice and are
recommended for the next generation of stock assessment models (Hoyle et
al. 2022; Punt 2023).

@@

As state-space models provide greater flexibility by allowing multiple
processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
al. 2018; Stock et al. 2021), one of the most immediate questions
regards the implications of mis-specification among alternative sources
of process error. Incorrect treatment of population attributes as
temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
could lead to misidentification of stock status and biased population
estimates, ultimately impacting fisheries management decisions (Legault
and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
Furthermore, biological, fishery, and observational processes are often
confounded in catch-at-age data, which may adversely affect the ability
to distinguish between true process variability and observational error
(Punt et al. 2014; Stewart and Monnahan 2017; Cronin-Fine and Punt 2021;
Fisch et al. 2023; Li et al. 2025a).

@@

Despite increasing research on state space assessment models, several
uncertainties in state space assessment modeling remain. First,
confounding processes that can be treated as random effects in the model
have not been thoroughly examined or tested within a
simulation-estimation framework. Second, previous studies relied on
operating models conditioned on specific fisheries, limiting their
generalizability (Liljestrand et al. 2024; Li et al. 2025a). In
particular, the effects of observation error and underlying fishing
history have not been fully isolated in simulation study designs, making
it challenging to disentangle the interplay between process and
observation error magnitudes, as demonstrated in Fisch et al. (2023).
Third, explicitly modeling stock-recruit relationships (SRRs) as
mechanistic drivers of population dynamics is promising (Fleischman et
al. 2013; Pontavice et al. 2022), but reliability of inferences within
integrated state-space age-structured models has not been evaluated.
Evidence from other studies suggests that when both process and
observation errors are unknown, estimating density dependence parameters
becomes highly uncertain (Knape 2008; Polansky et al. 2009). In
particular, Knape (2008) demonstrated that stronger density dependence
becomes increasingly difficult to estimate in the presence of
observation error. Therefore, it is crucial to assess whether
density-dependent mechanisms can be estimated with sufficient precision
for use in fisheries management (Auger-Méthé et al. 2016). Finally,
although the importance of autocorrelation in process errors is
recognized, investigations of the ability to distinguish state-space
assessment models with and without autocorrelation and whether such
misspecification is detrimental to estimation of important population
metrics are lacking (Johnson et al. 2016; Xu et al. 2019).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether an SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
whether OM and EM attributes affect rates of convergence and the ability
of Akaike Information Criterion (AIC) to correctly determine the source
of process error or the existence of an SRR. We also evaluate effects of
OM and EM attributes on magnitude of retrospective patterns and bias in
estimation of parameters and other model outputs important for
management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Stock and Miller 2021; Miller et al. 2025).
WHAM is an R package freely available via a Github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous Northeast United States (NEUS) fish stocks
(e.g., NEFSC 2022a, 2022b; NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed. R
OMs assume process error for recruitment only. Other OM categories
assume recruitment process errors along with process errors for apparent
survival (R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the R+S OMs as modeling apparent
survival because on log-scale the random effects are additive to the
total mortality (fishing and natural mortality) between numbers at age,
thus they modify the survival term. For each OM, assumptions about the
magnitude of the variance of process errors and observations are
required and the values we used were based on a review of the range of
estimates from NEUS assessments using WHAM.

In total, we configured 72 OMs with alternative assumptions about the
source and magnitude of process errors, magnitude of observation error
in indices and age composition data, and contrast in fishing pressure
over time. For each OM, we simulated 100 time series of abundance at age
with process errors, and for each realized time series, we simulated
observation data sets. For each data set, we fitted a number of EMs that
differed in assumptions about the source of process errors, whether
natural mortality (or the median for models with process error in
natural mortality) was estimated, and whether a Beverton-Holt SRR was
estimated within the EM. Details of each of the OMs and EMs are
described below. We did not use the log-normal bias-correction feature
for process errors or observations described by Stock and Miller (2021)
for OMs and EMs to simplify interpretation of the study results (Li et
al. 2025b). All code we used to perform the simulation study and
summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

We intended the population demographics and observation types to
represent a general NEUS groundfish stock. The population consists of 10
age classes, ages 1 to 10+, with the last being a plus group that
accumulates ages 10 and older. The maturity at age was a logistic curve
with \(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
left). Weight at age (\(W_a\)) was generated with a von Bertalanffy
growth function defining length at age: \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right),
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a
length-weight relationship such that \[
W_a = \theta_1 L_a^{\theta_2},
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\) (Figure
\ref{om_inputs_fig}, top right).

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. We assume spawning occurs annually 0.25 of each
year and recruitment at age 1 (\(N_{1,y}\)). All biological inputs to
calculations of spawning stock biomass (SSB) per recruit (i.e., weight,
maturity, and natural mortality at age) are constant in the R+S, R+Sel,
and R+q process error OMs. Therefore, steepness and unfished recruitment
are also constant over the time period for those OMs (Miller and Brooks
2021). We assumed a value of 0.2 for the natural mortality rate in OMs
without process errors on natural mortality. We specified unfished
recruitment equal to \(e^{10}\) and \(\Fmsy = F_{40\%} = 0.348\), which
equates to a steepness of 0.69 and \(a=0.60\) and
\(b = 2.4 \times 10^{-5}\) for the Beverton-Holt parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). For OMs with time-varying
random effects for natural mortality, steepness is not constant.
However, we used the same \(a\) and \(b\) parameters as other OMs, which
equates to a steepness and R0 at the median of the time series process
for natural mortality. Similarly, for OMs with time-varying random
effects for fishery selectivity, \Fmsy also varies temporally, so
equilibrium conditions for these OMs are defined for mean selectivity
parameters.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years followed by
fishing at \Fmsy for the last 20 years (denoted
\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
fished at \Fmsy for the entire time period (40 years). The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
the approach in Legault et al. (2023). The second scenario represents
the ideal situation where the stock is fished at an optimal level, but
provides less contrast in stock sizes over time. We specified initial
population abundance at age at the equilibrium distribution that
corresponds to fishing at either \(F = 2.5\Fmsy\) or \(F = \Fmsy\). This
implies that, for a deterministic model, the abundance at age would not
change from year to year at the beginning of the time series.

@@

For operating models with random effects on recruitment only and also on
apparent survival (R, R+S), we assumed marginal standard deviations for
recruitment of \(\sigma_R \in \{0.5,1.5\}\). The marginal standard
deviations for apparent survival random effects at older age classes
were \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial combination
of these process error assumptions (\(2\times 3\) levels) and scenarios
for fishing history (2 levels) and observation error (2 levels)
scenarios described above results in 24 different R
(\(\sigma_{2+} = 0\)) and R+S operating models (Table
\ref{naa_om_table}).

@@

WHAM treats each selectivity parameter \(s\) as a logit-transformed
parameter \[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\), where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters (\(a_{50}\) and slope
parameters) were bounded by \(s_l = 0\) and \(s_u = 10\) for all OMs and
EMs. The marginal standard deviations we assumed for logit scale random
effects were \(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation
parameters of \(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial
combination of these process error assumptions (2x2 levels) and
scenarios described above for fishing history (2 levels) and observation
error (2 levels) results in 16 different R+Sel OMs (Table
\ref{sel_om_table}).

@@

We investigated the reliability of AIC-based model selection for two
purposes. First, we analyzed selection of each process error model
source (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
simulated data set, we compared AIC for EMs with different process error
assumption conditional on whether median natural mortality rate and the
Beverton-Holt SRR were estimated. Second, we analyzed AIC-based
selection between EMs with and without the Beverton-Holt SRR assumed.
Contrast in fishing pressure and time series with recruitment at low
stock size have been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections indicated generally poor performance of AIC in determining
the Beverton-Holt SRR model for a given set of OM factors (including
contrast in fishing pressure), even when the EM was configured with the
correct process error source. Therefore, we conditioned on the EMs
having the correct process error assumption and also considered the
effect of the log-standard deviation of the true log(SSB)
(\(\log \text{SD}_{\text{SSB}}\); similar to the log of the coefficient
of variation for SSB) on model selection since simulations with realized
SSB producing low and high recruitment would have larger variation in
realized SSB.

@@

We also investigated bias in estimation of various model attributes as a
measure of reliability. For a given model attribute we calculated the
relative error \begin{equation}\label{relerror}
\text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
\end{equation} from fitting a given EM to simulated data set \(j\)
configured for a given OM where \(\widehat \theta_j\) and \(\theta_j\)
are the estimated and true values for simulation \(j\). We analyzed
simulation results for estimates of terminal year SSB and recruitment,
Beverton-Holt SRR parameters (\(a\) and \(b\)), and median natural
mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted
models as a measure of reliability. We calculated Mohn's \(\rho\) for
SSB, fishing mortality (averaged over all age classes), and recruitment
for each EM fit to each OM simulated data set (Mohn 1999). We fit
\(P = 7\) peels to each simulated data set and calculated Mohn's
\(\rho\) for a given attribute \(\theta\) as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(Y\) is last year of the full set of observations
and \(\widehat \theta_{y,y'}\) is the estimate for attribute \(\theta\)
in year \(y\) from a model fit using data up to year \(y'\geq y\). Thus,
\(\theta\) terms where \(y'=Y\) refer to estimates from the fit to all
years of data.

@@

Because the OM and EM attributes that we investigated are numerous, we
used two methods to summarize the most important factors for differences
in results within a given OM process error source. The first method was
fitting regression models with the response being each of the measures
of reliability described above and predictor variables were defined
based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
al. 2017; Harwell et al. 2018). For the binary indicators of convergence
and AIC-based selection of an SRR, we performed logistic regressions.
For indicators of AIC-based selection of EM process error source
(multiple categories) we performed multinomial regressions. For other
measures of reliability we fit linear regression models to transformed
responses. Because relative errors (Eq. \ref{relerror}) and Mohn's
\(\rho\) for the various parameters are bounded below at -1, we used a
transformation of these values
\begin{equation}\label{bias_regression_response}
y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(j\), so that values are unbounded. For relative errors, \(y_j\) is the
log-scale error. We omitted simulations where estimated attributes equal
to zero (RE = -1). For all regressions we fit separate models with just
individual OM and EM factors included, with all factors included, with
all second order interactions, and with all third order interactions.
For the multinomial regression, we used the \verb|vglm| function from
the VGAM package (Yee 2008; Yee 2015). We tabulated percent reduction in
residual deviance for each of the regression fits. We did not perform
formal statistical analyses of effects of OM and EM attributes on
results (e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

The second method involved fitting classification and regression trees
(Breiman et al. 1984) to show how the OM and EM attributes, and their
interactions, partition the values for each measure of reliability
(e.g., Gonzalez et al. 2018; Collier et al. 2022). We used
classification trees for categorical measures (convergence and AIC) and
regression trees for the other measures with continuous scales (relative
error and Mohn's \(\rho\)). The response variables were the same as the
regressions for the deviance reduction analyses. We used the
\verb|rpart| function in the rpart package (Therneau and Atkinson 2025)
to fit trees. Full trees were determined using default settings except
that we increased the number of cross-validations to 100. For clarity,
we manually pruned the full trees to show just the primary branches.

We also provide detailed results for all measures of reliability at each
combination of OM and EM attributes in the Supplementary Materials. For
confidence intervals of probability of convergence, we used the
Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
For AIC selection of process error source we provide estimates of the
proportions of simulations where each EM type was selected. For AIC
selection of the SRR (a binary indicator for each simulated data set),
we fit logistic regressions and present resulting predicted
probabilities of correctly selecting the SRR as a function of SSB
variability (\(\log \text{SD}_{\text{SSB}}\) described above). We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias, and Mohn's \(\rho\)
using the binomial distribution approach (Thompson 1936) as in Miller
and Hyun (2018) and Stock and Miller (2021).
